The use of lithic assemblages for the definition of short-term occupations in hunter-gatherer prehistory

### 2018-04-25 analysis/paper/paper.Rmd
paper submitted

---
title: "The use of lithic assemblages for the definition of short-term occupations in hunter-gatherer prehistory"
author:
  - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal) nbicho@ualg.pt
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
date: '`r Sys.Date()`'
output: 
    bookdown::word_document2: #subsistute by pdf_document2 or hmtl_document2 as necessary
      fig_caption: yes
      reference_docx: "../templates/template2.docx" 
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl"
abstract: |
  One of the main elements in prehistoric research is the study of settlement patterns. In the last five decades, stemming partially from Binford’s research on the topic, the idea of settlement patterns is based on site typology, including the traditional residential and logistic concepts. Both models of land use and exploitation are certainly marked by the notion of short-term occupation. This concept, used freely by many archaeologists, tends to rely on two main ideas: an occupation lasted a short span of time and resulted on a limited amount of material culture.Our aim, based on our results from various archaeological case studies dated to the Stone Age records of Portugal, is to show that neither idea is necessarily correct: e.g. there may be short-term occupations with the production of large amounts of artifacts such as lithic workshops; there might be very small collections, such as lithic caches, resulting from short occupations but with very long uses of the site; and most times both are hardly differentiated within complex palimpsests.Our study shows that the common use of lithic volumetric density and retouch frequency is not always sufficient to differentiate between short and long-term occupations. Also, there are other variables that are more sensitive to indicate the duration of occupation of an archaeological context that should be used identification of time length.

@@

Prehistoric archaeology has been fighting hard to understand the archaeological record and how it relates to past human evolution and anthropogenic adaptations to the changing environment. The development of actualistic studies (sensu Binford 1981) in the last half century, including such specialized disciplines as Ethnoarchaeology, Geoarchaeology or Zooarchaeology, have greatly helped to understand not only the archaeological material culture but also site formation processes (e.g., Schiffer 1983, 1987) in a wide and very diversified manner. Nevertheless, archaeologists still endure complex problems for the definition of universally used concepts related to past human adaptations. In some cases, however, archaeologists just freely use those concepts without the necessary careful or proper consideration of its meaning and the impact that they might have in further interpretations of the distant past. The concept of ‘short-term occupation’ seems to be such a case. The idea of short-term occupation has been frequently used to succinctly describe site´s or lithic assemblage’s characteristics (e.g., Porraz 2009; Picin 2017; Rios-Garaizar 2016). Most times, it is, in one way or another, related to work on settlement patterns, starting with the classic Binford’ studies on the Nunamiut settlement system and related discussion on site structuring and intra-site spatial organization (e.g., Binford 1982, 1978a, 1978b), as well as on hunter-gatherer mobility research (e.g., Amick 1996; Binford 1980; Kelly 1995; Kuhn 1992; Nishiaki and Akazawa 2015; Barton and Riel-Salvatore 2014). 
In 1972, Sahlins argued that, if given enough time, a small group of foragers can rapidly deplete the resources within a short range from their residential base camp. This basic assumption is fundamental to understand the value of the definition of a short-term occupation for hunter-gatherer research, and more so in the case of the prehistoric archaeological record. As Yellen (1977) stated, the duration of occupation of a group in a specific ecological context will affect both the quantity and nature of the archaeological record visible at a site. The issue, then, is to jiggle a series of variables, including duration of occupation and population size, and to make sense of the results, so the concept is clear and used in a meaningful manner. 
As Moncel and Rivals (2011) have argued, it is not always easy to make a distinction between long-term and short-term occupations. While ethnographically this distinction might be easy to demonstrate, although most certainly corresponding to a continuum as both Yellen (1977) and Binford (1978a) have shown in their ethnoarchaeological studies, in the archaeological record this task is certainly more complex. There are a set of cultural variables that have a direct impact on the duration of any occupation. The most important are likely to be population size, habitat quality (or in other words, local ecological diversity and carrying capacity), diversity and type of functions at the site. These, unfortunately, cannot be measured directly in the archaeological record and we need proxies that helps us to understand the diversity of each at every single site, and even in individual contexts at each site. While in the case of population size there has been attempts to produce numerical formulae to resolve the issue (e.g., Grove 2009), the most common variable used to define the size of a population at a site is the dimension of the site. The measurement of the habitat quality is certainly more problematic and is limited to organic data (i.e., fauna and flora in all possible formats) and raw materials, present at the site. Site diversity and type of functions have traditionally been measured based on, respectively, tool diversity indices, and use-wear and residue analyses. In addition to tool diversity, the presence, quantity and diversity of habitat features is also commonly used for determining site function (Table \@ref(tab:table1)).

```{r table1}

#read table
table1 <- readr::read_csv("../data/raw_data/traditional_criteria.csv")

#show table
knitr::kable(table1,
             digits = 2,
             row.names = FALSE,
             caption= "Traditional criteria for estimation of duration of occupation"
             )

@@

The use of the concept of short-term (and also that of long-term) occupations is directly related, in the archaeological literature, with site typology. Terms such as Residential site, Residential camp, Logistical camp, Base camp, and many others are frequently applied with no particular verified criteria other than the size of the site, or of the dimension and diversity of the artefactual assemblage (e.g., Bretzke et al. 2017; Crassard et al. 2013; Madsen et al. 2006; Porraz 2009; Rios-Garaizar 2016; Terradillos-Bernal et al. 2017). Other studies have furthered improved those concepts, with both descriptions of criteria and their application (e.g., Barton 1990; Clark and Barton 2017; Dillehay et al. 2011; Nishiaki and Akazawa 2015). 
In Binford’s seminal framework there are clear definitions of site typology (Binford 1980-12) that relate to mobility patterns:
“For foragers, I recognized two types of site, the residential base camp and the location. Collectors generate at least three additional types of sites by virtue of the logistical character of their procurement strategies. These I have designated the field camp, the station, and the cache.” (Binford 1980:10).
Another good example of site nomenclature is that of Dillehay et al. (2011) for the foraging-farming transition in the Andes. The authors indicate the presence of a very diverse and expanded site typology (Dillehay et al. 2011:36-40), defining long-term and short-term base and field camps, processing stations, transitory station/workshops, lithic quarries, earthen mounds, horticultural residences with gardens, permanent residences associated with irrigation, agriculture, hillside villages and special activity locales. 
These examples start with a basic assumption that “…the greater the number of generic types of functions a site may serve, the greater the number of possible combinations, and hence the greater the range of inter-site variability…” (Binford 1980:12). However, as Binford also argued, there is considerable variability in the duration of each stay in each site and the type of mobility pattern present (Binford 1980), and thus each site may have been used for different functions and durations (Binford 1982), with a direct impact on the archaeological visibility of sites and their respective duration cycle. Barton and colleagues (Barton and Riel-Salvatore 2014; Clark and Barton 2017; Riel-Salvatore and Barton 2004), following Binford’s perspective, have argued for a continuum in the use of space and time between residential (Foragers) and logistical (Collectors) patterns, that can be measured in the archaeological record using the relationship between artifact volumetric density and the frequency of retouched tools within each assemblage – named by the authors as the Whole Assemblage Behavioral Index (WABI). The index is expected to show a negative correlation between both variables, reflecting accumulated artifacts deriving from primarily curated to primarily expedient artifact use.  Although the authors emphasize that “the terms “expedient” and “curated” these do not reflect individual site-occupation events” (Riel-Salvatore and Barton, 2007, p. 62). They also assume that expedient assemblages often accumulate at more intensively occupied sites, while curated assemblages more usually derive from short-term occupations.
In this study, we focus on previous uses of the term short-term occupations and its definitions, and the parent archaeological proxies used to define length of site occupation as well as site type. Using data from various case studies we have been working with over the past 25 years, we critically evaluate the application of the WABI proxy and explore a combination of other potential variables for the definition of occupation duration using commonly available stone tool and context data.

@@

We used in this study a total of 17 different archaeological contexts, coming from a set of various sites excavated since the late 1980’s: Areeiro I (Bicho 1992, 1993, 1994) Cabeço do Porto Marinho (Bicho 1992, 1994), Carneira II  (Bicho 1992, 1993), Picareiro (Bicho et al. 2006), Pinhal da Carneira (Bicho 1992, 1993), and Quinta do Sanguinhal (Bicho 2005) in central Portugal, and Vale Boi (Bicho et al. 2017b; Cascalheira and Bicho 2013, 2015) in southern Portugal. 
Areeiro I (ARI) was excavated in 1987 under Anthony Marks direction (Marks et al. 1994). It is located in the Rio Maior area, in a region with the highest concentration of open-air sites in the Portuguese Upper Paleolithic.  It was found in sands, as all the Rio Maior sites in the present paper. Excavation was carried out by artificial 10 cm spits and all sediment was screened through a 2 mm mesh. Artifacts are slightly damaged by a modern fire. No habitat features were found at the excavated area (Table \@ref(tab:table2)), but the density of artifacts and the apparent extension of the deposit suggested a medium to long-term residential occupation.

```{r table2}

#Load dataset
Dataset <- readr::read_csv("../data/raw_data/raw_dataset.csv")

#Calculate frequency variables
Dataset$LithicDensity <- Dataset$Artifacts / Dataset$SampledVolume
Dataset$CoreFreq <- Dataset$Cores / Dataset$Artifacts
Dataset$BlanksFreq <- Dataset$Blanks / Dataset$Artifacts
Dataset$ChipsFreq <- Dataset$Chips / Dataset$Artifacts
Dataset$RetouchFreq <- Dataset$RetouchedTools / Dataset$Artifacts
Dataset$ToolDiversity <- Dataset$ToolTypes / (sqrt(Dataset$RetouchedTools))
Dataset$FeaturesFreq <- Dataset$Features / Dataset$SampledVolume

#Show Table 2
knitr::kable(Dataset,
             digits = 2,
             row.names = FALSE,
             caption = "Archaeological data based on variables from Table 1"
             )

@@

Cabeço do Porto Marinho (CPM) was excavated between 1987 and 1994 (Bicho 1992; Marks et al. 1994). CPM is one of the largest Portuguese Upper Paleolithic sites with a very long archaeological sequence. There are over 30 different stratigraphical contexts starting with early Gravettian, Proto-Solutrean, Magdalenian, Epipaleolithic, Neolithic and Bronze Age occupations (Bicho 1992; Marks et al. 1994; Zilhão 1997). For this study we used various horizons, known as CPM I Lower, I Upper, II Middle, II Upper, III Upper, III South, III Trench, and V, respectively coming from loci I, II, III and V. All assemblages are dated to the Magdalenian and Epipaleolithic, between c. 20 and 9 ka cal BP. Artifact retrieval resulted from excavation based on artificial 5 or 10 cm spits and all sediment was screened with a 2 mm mesh, but only CMP III trench was fully excavated. Some of those contexts had at least an in situ hearth (Bicho 1992). Data from each context is presented in Table \@ref(tab:table2), but the lithic assemblages have between c. 1500 artifacts (CPM III Trench) and >10 000 artifacts (CPM I Upper), although only the samples studied by Bicho (1992) were used in the present study. With the exception of CPM III Trench, corresponding to a small area around a single hearth, all other contexts were thought to be of middle to long-term occupations, likely residential base camps.
Carneira II (CR II) and Pinhal da Carneira (PC) are sites in the same general pine grove outside Rio Maior city. They are both single layered sites with an average of 20-25 cm thick deposit. There is a heart in CR II, but no features were found in the small excavated area of PC. Excavation was carried out in 10 cm artificial spits and all sediment was screened with a 2 mm mesh. Both sites are dated to the Epipaleolithic, respectively c. 10.5 and 11.5 cal BP (Bicho 1992, 1994). Lithic assemblages used in this study are samples from the excavated area (Table \@ref(tab:table2)). Both horizons, based on the amount of artifacts, were thought to be middle to long-term occupations, likely residential base camps.
Quinta do Sanguinhal (QS) is a very small open air site, also in the Rio Maior region (Bicho 2005). It is a Gravettian occupation, fully excavated with a total of 6 m2 (although it may have been slightly larger but, when we found the site, the western section had been removed due to construction of a building). The assemblage is composed of c. 1500 artifacts (Table \@ref(tab:table2)) and there are several refittings. No features were found in the excavated area. Based on the lithic assemblage, the occupation was thought to correspond to a single short-term blade production site, with very few tools and cores, with many blades missing from the sequence, as shown by reffiting(Almeida 2000; Bicho 2005).
The cave site of Picareiro is located in central Portugal. The cave was excavated between 1994 and 2001 (PI Nuno Bicho) and a second project started in 2005 and is still underway under the direction of Jonathan Haws. The cave has a very long sequence of more than 9 meters and has archaeological horizons dated from the Middle Paleolithic to the Bronze Age (Bicho et al. 2006; Haws 2012). The lithic assemblage used here dates to the Magdalenian and is associated to a very large hearth (Bicho et al. 2006). This specific occupation was thought to correspond to a meat drying and smoking logistical station. It seems to be a very specialized industry composed mostly of chipage, a few cores, and small backed bladelets (both points such as microgravettes, and simple backed pieces). The sample is close to 2000 artifacts (Table \@ref(tab:table2)).
The site of Vale Boi was discovered in 1998 and excavation started in 2000 (Bicho et al. 2004; Bicho et al. 2013). There are four different loci (Terrace, Slope, Rock Shelter and Rock Shelter 2) with archaeological horizons dated to the early Gravettian (Bicho et al. 2017a; Bicho et al. 2015), Solutrean and Magdalenian, while one of those loci (VB Terrace) has also Mesolithic and early Neolithic occupations (Bicho 2009; Carvalho 2007). Excavation followed detailed 3D location of artifacts using a Total station and full sediment screening with a 2-3 mm mesh screen.
The oldest assemblage is early Gravettian. It is a particular setting found at the bottom of VB Rock Shelter 2, composed by only 54 artifacts and coming from an area slightly smaller than 2 m2. It likely corresponds to a Gravettian point cache (Bicho et al. 2016), since the large majority of the recovered artifacts are backed pointed bladelets.
The other two assemblages (VB Shelter B and Z) are from the rock shelter locus, respectively dated to the Solutrean and Magdalenian (Cascalheira 2013; Cascalheira and Bicho 2015; Mendonça 2009). No habitat features were found in the small Magdalenian horizon but a small hearth was found in the middle of the excavated area of the VB B Solutrean Layer (Table \@ref(tab:table2)).

@@

The literature referenced before indicates as the main archeological variables for the definition of the duration of occupation, artifact and tool density (Barton and Riel-Salvatore 2014; Clark and Barton 2017), expressed as retouch frequency and lithic volumetric density (artifacts per cubic meter). Logistically base camps and short-term camps were recognized by Clark and Barton (2017) based on the correlation of those two variables, producing a table of reference of material correlates of mobility. Short-term camps represent overnight and limited activities of small groups out from logistically base camps or larger groups based on a regular residential moving pattern. The result is a high incidence of relative retouch but with low lithic volumetric density, high lithic curation, low numbers of cores and debitage, and small archaeological contexts. By opposition, sites with large residential stability, named by those authors as logistically organized base camps, are marked by high lithic densities, low incidence of retouch, high numbers of non-exhausted cores, high numbers of blanks, large sites and what are usually called expedient technologies and assemblages.
There are however, other variables that have been listed as possible indicative of duration of occupation. These include site area, number of features, and spatial segregation of activities (e.g., Binford 1980; Dillehay et al. 2011; Nishiaki and Akazawa 2015). 
In the present study we use all of the above variables, so we can test their usefulness as a measure of time of occupation in archaeological contexts:

@@

*	**Lithic Density** - estimated number of artifacts present in one cubic meter of sediment (following Clark and Barton 2017);

@@

*	**Retouch Frequency** - relative frequency of retouched artifacts in the lithic sample (following Clark and Barton 2017);

@@

Principal Component Analysis (PCA) is a commonly used technique to extract relevant information from a multivariate dataset and to express this information as a set of few new variables called principal components or dimensions. The usefulness of PCA is that, in a single analytical process, one can indicate relationships between and within variables and cases, suggest general trends in data structure and identify which variables best explain these patterns, compress large percentages of variance from a wide set of variables in a reduced number of factors, and perform this transformation so that the new variables are not correlated and therefore do not present redundant information (Shennan, 1988). PCA has had numerous applications in Archaeology (e.g. Lycett et al., 2006; Marreiros and Bicho, 2013; McCall 2007; 2006; McPherron, 1994) and is available through most of the statistical software packages.
Here, we applied PCA to our dataset to identify possible correlations between variables that would indicate the existence of patterns that, when compared with a priori information from each of the sites, could possibly be translated into more secure proxies for occupation duration.
All analyses and data processing were accomplished in R (version 3.3.1) (R Core Team, 2015). PCA analysis was performed using the FactoMineR package (Le et al., 2008). Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our supplemental online material (SOM) at https://dx.doi.org/… To produce those files we followed the procedures described by Marwick et al. (2017) for the creation of research compendiums to enhance the reproducibility of research. The files provided contain all the raw data used in our analysis as well as a custom R package (Wickham, 2015) holding the code use for all analysis and to produce all tables and figures. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, (for more information see Marwick, 2016).

@@

As a result, running PCA with the enlarged set of criteria provides a slightly different perspective on the WABI patterns. Four of the calculated PCA dimensions present eigenvalues higher than 1, explaining more than 87% of dataset variability (Table \@ref(tab:table3)). Dimensions 1 and 2, alone, explain over 58% of the variability.

@@

ShortTerm <- Dataset[vars]

# Set site names as row.names
row.names(ShortTerm) <- Dataset$Sites

@@

```{r table3}

>>>>

---
title: "The use of lithic assemblages for the definition of short-term occupations in hunter-gatherer prehistory"
author:
  - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal) nbicho@ualg.pt
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
date: '`r Sys.Date()`'
output: 
    bookdown::word_document2: #subsistute by pdf_document2 or hmtl_document2 as necessary
      fig_caption: yes
      reference_docx: "../templates/template2.docx" 
bibliography: ShortTermOccupations.bib
csl: "../templates/springer-humanities-author-date.csl"
abstract: |
  One of the main elements in prehistoric research is the study of settlement patterns. In the last five decades, stemming partially from Binford’s research on the topic, the idea of settlement patterns is based on site typology, including the traditional residential and logistic concepts. Both models of land use and exploitation are certainly marked by the notion of short-term occupation. This concept, used freely by many archaeologists, tends to rely on two main ideas: an occupation lasted a short span of time and resulted on a limited amount of material culture.Our aim, based on our results from various archaeological case studies dated to the Stone Age records of Portugal, is to show that neither idea is necessarily correct: e.g. there may be short-term occupations with the production of large amounts of artifacts such as lithic workshops; there might be very small collections, such as lithic caches, resulting from short occupations but with very long uses of the site; and most times both are hardly differentiated within complex palimpsests.Our study shows that the common use of lithic volumetric density and retouch frequency is not always sufficient to differentiate between short and long-term occupations. Also, there are other variables that are more sensitive to indicate the duration of occupation of an archaeological context that should be used identification of time length.

@@

Prehistoric archaeology has been fighting hard to understand the archaeological record and how it relates to past human evolution and anthropogenic adaptations to the changing environment. The development of actualistic studies [sensu @binford_bones:_1981] in the last half century, including such specialized disciplines as Ethnoarchaeology, Geoarchaeology or Zooarchaeology, have greatly helped to understand not only the archaeological material culture but also site formation processes [e.g. @schiffer_toward_1983; @schiffer_formation_1987] in a wide and very diversified manner. Nevertheless, archaeologists still endure complex problems for the definition of universally used concepts related to past human adaptations. In some cases, however, archaeologists just freely use those concepts without the necessary careful or proper consideration of its meaning and the impact that they might have in further interpretations of the distant past. The concept of ‘short-term occupation’ seems to be such a case. The idea of short-term occupation has been frequently used to succinctly describe site´s or lithic assemblage’s characteristics [e.g. @porraz_middle_2009; @picin_short-term_2017;  @rios-garaizar_early_2016]. Most times, it is, in one way or another, related to work on settlement patterns, starting with the classic Binford’ studies on the Nunamiut settlement system and related discussion on site structuring and intra-site spatial organization [e.g. @binford_dimensional_1978; @binford_nunamiut_1978; @binford_archaeology_1982], as well as on hunter-gatherer mobility research [e.g. @binford_willow_1980; @kuhn_planning_1992; @kelly_foraging_1995; @amick_regional_1996; @barton_formation_2014; @nishiaki_patterning_2015].

In 1972, @sahlins_stone_1972 argued that, if given enough time, a small group of foragers can rapidly deplete the resources within a short range from their residential base camp. This basic assumption is fundamental to understand the value of the definition of a short-term occupation for hunter-gatherer research, and more so in the case of the prehistoric archaeological record. As Yellen [-@yellen_archaeological_1977] stated, the duration of occupation of a group in a specific ecological context will affect both the quantity and nature of the archaeological record visible at a site. The issue, then, is to jiggle a series of variables, including duration of occupation and population size, and to make sense of the results, so the concept is clear and used in a meaningful manner.

As Moncel and Rivals [-@moncel_question_2011] have argued, it is not always easy to make a distinction between long-term and short-term occupations. While ethnographically this distinction might be easy to demonstrate, although most certainly corresponding to a continuum as both Yellen [-@yellen_archaeological_1977] and Binford [-@binford_dimensional_1978] have shown in their ethnoarchaeological studies, in the archaeological record this task is certainly more complex. There are a set of cultural variables that have a direct impact on the duration of any occupation. The most important are likely to be population size, habitat quality (or in other words, local ecological diversity and carrying capacity), diversity and type of functions at the site. These, unfortunately, cannot be measured directly in the archaeological record and we need proxies that helps us to understand the diversity of each at every single site, and even in individual contexts at each site. While in the case of population size there has been attempts to produce numerical formulae to resolve the issue [e.g. @grove_huntergatherer_2009], the most common variable used to define the size of a population at a site is the dimension of the site. The measurement of the habitat quality is certainly more problematic and is limited to organic data (i.e., fauna and flora in all possible formats) and raw materials, present at the site. Site diversity and type of functions have traditionally been measured based on, respectively, tool diversity indices, and use-wear and residue analyses. In addition to tool diversity, the presence, quantity and diversity of habitat features is also commonly used for determining site function (Table \@ref(tab:traditional)).

\newline

```{r traditional}

#read table
traditional <- readr::read_csv("../data/raw_data/traditional_criteria.csv")

#show table
knitr::kable(traditional,
             digits = 2,
             row.names = FALSE,
             caption= "Traditional criteria for estimation of duration of occupation"
             )

@@

\newline

The use of the concept of short-term (and also that of long-term) occupations is directly related, in the archaeological literature, with site typology. Terms such as Residential site, Residential camp, Logistical camp, Base camp, and many others are frequently applied with no particular verified criteria other than the size of the site, or of the dimension and diversity of the artefactual assemblage [e.g. @madsen_late_2006; @crassard_middle_2013; @bretzke_middle_2017; @porraz_middle_2009; @rios-garaizar_early_2016; @terradillos-bernal_san_2017]. Other studies have furthered improved those concepts, with both descriptions of criteria and their application [e.g. @barton_stone_1990; @clark_lithics_2017; @dillehay_research_2011; @nishiaki_patterning_2015].

In Binford’s seminal framework there are clear definitions of site typology [@binford_willow_1980] that relate to mobility patterns:
“For foragers, I recognized two types of site, the residential base camp and the location. Collectors generate at least three additional types of sites by virtue of the logistical character of their procurement strategies. These I have designated the field camp, the station, and the cache.” [@binford_willow_1980 :10].

Another good example of site nomenclature is that of Dillehay et al. [-@dillehay_research_2011] for the foraging-farming transition in the Andes. The authors indicate the presence of a very diverse and expanded site typology [@dillehay_research_2011 :36-40], defining long-term and short-term base and field camps, processing stations, transitory station/workshops, lithic quarries, earthen mounds, horticultural residences with gardens, permanent residences associated with irrigation, agriculture, hillside villages and special activity locales.

These examples start with a basic assumption that “…the greater the number of generic types of functions a site may serve, the greater the number of possible combinations, and hence the greater the range of inter-site variability…” [@binford_willow_1980 :12]. However, as Binford also argued, there is considerable variability in the duration of each stay in each site and the type of mobility pattern present [@binford_willow_1980], and thus each site may have been used for different functions and durations [@binford_archaeology_1982], with a direct impact on the archaeological visibility of sites and their respective duration cycle. Barton and colleagues [@riel-salvatore_late_2004; @barton_formation_2014; @clark_lithics_2017], following Binford’s perspective, have argued for a continuum in the use of space and time between residential (Foragers) and logistical (Collectors) patterns, that can be measured in the archaeological record using the relationship between artifact volumetric density and the frequency of retouched tools within each assemblage – named by the authors as the Whole Assemblage Behavioral Index (WABI). The index is expected to show a negative correlation between both variables, reflecting accumulated artifacts deriving from primarily curated to primarily expedient artifact use.  Although the authors emphasize that “the terms “expedient” and “curated” these do not reflect individual site-occupation events” [@riel-salvatore_new_2007 :62]. They also assume that expedient assemblages often accumulate at more intensively occupied sites, while curated assemblages more usually derive from short-term occupations.

In this study, we focus on previous uses of the term short-term occupations and its definitions, and the parent archaeological proxies used to define length of site occupation as well as site type. Using data from various case studies we have been working with over the past 25 years, we critically evaluate the application of the WABI proxy and explore a combination of other potential variables for the definition of occupation duration using commonly available stone tool and context data.

@@

We used in this study a total of 17 different archaeological contexts, coming from a set of various sites excavated since the late 1980’s: Areeiro I [@bicho_end_1994; @bicho_late_1993; @bicho_technological_1992] Cabeço do Porto Marinho [@bicho_end_1994; @bicho_technological_1992], Carneira II [@bicho_late_1993; @bicho_technological_1992], Picareiro [@bicho_two_2006], Pinhal da Carneira [@bicho_late_1993; @bicho_technological_1992], and Quinta do Sanguinhal [@bicho_extinction_2005] in central Portugal, and Vale Boi [@bicho_rapid_2017; @cascalheira_chronological_2015; @cascalheira_huntergatherer_2013] in southern Portugal.

Areeiro I (ARI) was excavated in 1987 under Anthony Marks direction [@marks_upper_1994]. It is located in the Rio Maior area, in a region with the highest concentration of open-air sites in the Portuguese Upper Paleolithic.  It was found in sands, as all the Rio Maior sites in the present paper. Excavation was carried out by artificial 10 cm spits and all sediment was screened through a 2 mm mesh. Artifacts are slightly damaged by a modern fire. No habitat features were found at the excavated area (Table \@ref(tab:dataset)), but the density of artifacts and the apparent extension of the deposit suggested a medium to long-term residential occupation.

\newline

```{r dataset}

#Load dataset
dataset <- readr::read_csv("../data/raw_data/raw_dataset.csv")

#Calculate frequency variables
dataset$LithicDensity <- dataset$Artifacts / dataset$SampledVolume
dataset$CoreFreq <- dataset$Cores / dataset$Artifacts
dataset$BlanksFreq <- dataset$Blanks / dataset$Artifacts
dataset$ChipsFreq <- dataset$Chips / dataset$Artifacts
dataset$RetouchFreq <- dataset$RetouchedTools / dataset$Artifacts
dataset$ToolDiversity <- dataset$ToolTypes / (sqrt(dataset$RetouchedTools))
dataset$FeaturesFreq <- dataset$Features / dataset$SampledVolume

#Show Table 2
knitr::kable(dataset,
             digits = 2,
             row.names = FALSE,
             caption = "Archaeological data based on variables from Table 1"
             )

@@

\newline

Cabeço do Porto Marinho (CPM) was excavated between 1987 and 1994 [@bicho_technological_1992; @marks_upper_1994]. CPM is one of the largest Portuguese Upper Paleolithic sites with a very long archaeological sequence. There are over 30 different stratigraphical contexts starting with early Gravettian, Proto-Solutrean, Magdalenian, Epipaleolithic, Neolithic and Bronze Age occupations [@bicho_technological_1992; @marks_upper_1994; @zilhao_o_1997]. For this study we used various horizons, known as CPM I Lower, I Upper, II Middle, II Upper, III Upper, III South, III Trench, and V, respectively coming from loci I, II, III and V. All assemblages are dated to the Magdalenian and Epipaleolithic, between c. 20 and 9 ka cal BP. Artifact retrieval resulted from excavation based on artificial 5 or 10 cm spits and all sediment was screened with a 2 mm mesh, but only CMP III trench was fully excavated. Some of those contexts had at least an in situ hearth [@bicho_technological_1992]. Data from each context is presented in Table \@ref(tab:dataset), but the lithic assemblages have between c. 1500 artifacts (CPM III Trench) and >10 000 artifacts (CPM I Upper), although only the samples studied by Bicho [-@bicho_technological_1992] were used in the present study. With the exception of CPM III Trench, corresponding to a small area around a single hearth, all other contexts were thought to be of middle to long-term occupations, likely residential base camps.

Carneira II (CR II) and Pinhal da Carneira (PC) are sites in the same general pine grove outside Rio Maior city. They are both single layered sites with an average of 20-25 cm thick deposit. There is a heart in CR II, but no features were found in the small excavated area of PC. Excavation was carried out in 10 cm artificial spits and all sediment was screened with a 2 mm mesh. Both sites are dated to the Epipaleolithic, respectively c. 10.5 and 11.5 cal BP [@bicho_end_1994; @bicho_technological_1992]. Lithic assemblages used in this study are samples from the excavated area (Table \@ref(tab:dataset)). Both horizons, based on the amount of artifacts, were thought to be middle to long-term occupations, likely residential base camps.

Quinta do Sanguinhal (QS) is a very small open air site, also in the Rio Maior region [@bicho_extinction_2005]. It is a Gravettian occupation, fully excavated with a total of 6 m2 (although it may have been slightly larger but, when we found the site, the western section had been removed due to construction of a building). The assemblage is composed of c. 1500 artifacts (Table \@ref(tab:dataset)) and there are several refittings. No features were found in the excavated area. Based on the lithic assemblage, the occupation was thought to correspond to a single short-term blade production site, with very few tools and cores, with many blades missing from the sequence, as shown by reffiting [@almeida_terminal_2000; @bicho_extinction_2005].

The cave site of Picareiro is located in central Portugal. The cave was excavated between 1994 and 2001 (PI Nuno Bicho) and a second project started in 2005 and is still underway under the direction of Jonathan Haws. The cave has a very long sequence of more than 9 meters and has archaeological horizons dated from the Middle Paleolithic to the Bronze Age [@bicho_two_2006; @haws_paleolithic_2012]. The lithic assemblage used here dates to the Magdalenian and is associated to a very large hearth [@bicho_two_2006]. This specific occupation was thought to correspond to a meat drying and smoking logistical station. It seems to be a very specialized industry composed mostly of chipage, a few cores, and small backed bladelets (both points such as microgravettes, and simple backed pieces). The sample is close to 2000 artifacts (Table \@ref(tab:dataset)).

The site of Vale Boi was discovered in 1998 and excavation started in 2000 [@bicho_ecodynamics_2013; @bicho_shell_2004]. There are four different loci (Terrace, Slope, Rock Shelter and Rock Shelter 2) with archaeological horizons dated to the early Gravettian [@bicho_bayesian_2015; @bicho_early_2017], Solutrean and Magdalenian, while one of those loci (VB Terrace) has also Mesolithic and early Neolithic occupations [@bicho_edge:_2009; @carvalho_neolitizacao_2007]. Excavation followed detailed 3D location of artifacts using a Total station and full sediment screening with a 2-3 mm mesh screen.

The oldest assemblage is early Gravettian. It is a particular setting found at the bottom of VB Rock Shelter 2, composed by only 54 artifacts and coming from an area slightly smaller than 2 m2. It likely corresponds to a Gravettian point cache [@bicho_early_2016], since the large majority of the recovered artifacts are backed pointed bladelets.

The other two assemblages (VB Shelter B and Z) are from the rock shelter locus, respectively dated to the Solutrean and Magdalenian [@cascalheira_chronological_2015; @cascalheira_influencia_2013; @mendonca_tecnologia_2009]. No habitat features were found in the small Magdalenian horizon but a small hearth was found in the middle of the excavated area of the VB B Solutrean Layer (Table \@ref(tab:dataset)).

@@

The literature referenced before indicates as the main archeological variables for the definition of the duration of occupation, artifact and tool density (Barton and Riel-Salvatore 2014; Clark and Barton 2017), expressed as retouch frequency and lithic volumetric density (artifacts per cubic meter). Logistically base camps and short-term camps were recognized by Clark and Barton [-@clark_lithics_2017] based on the correlation of those two variables, producing a table of reference of material correlates of mobility. Short-term camps represent overnight and limited activities of small groups out from logistically base camps or larger groups based on a regular residential moving pattern. The result is a high incidence of relative retouch but with low lithic volumetric density, high lithic curation, low numbers of cores and debitage, and small archaeological contexts. By opposition, sites with large residential stability, named by those authors as logistically organized base camps, are marked by high lithic densities, low incidence of retouch, high numbers of non-exhausted cores, high numbers of blanks, large sites and what are usually called expedient technologies and assemblages.

There are, however, other variables that have been listed as possible indicative of duration of occupation. These include site area, number of features, and spatial segregation of activities [e.g. @binford_willow_1980; @dillehay_research_2011; @nishiaki_patterning_2015]. 
In the present study we use all of the above variables, so we can test their usefulness as a measure of time of occupation in archaeological contexts:

@@

*	**Lithic Density** - estimated number of artifacts present in one cubic meter of sediment (following Clark and Barton [-@clark_lithics_2017]);

@@

*	**Retouch Frequency** - relative frequency of retouched artifacts in the lithic sample (following Clark and Barton [-@clark_lithics_2017]);

@@

Principal Component Analysis (PCA) is a commonly used technique to extract relevant information from a multivariate dataset and to express this information as a set of few new variables called principal components or dimensions. The usefulness of PCA is that, in a single analytical process, one can indicate relationships between and within variables and cases, suggest general trends in data structure and identify which variables best explain these patterns, compress large percentages of variance from a wide set of variables in a reduced number of factors, and perform this transformation so that the new variables are not correlated and therefore do not present redundant information [@shennan_quantifying_1997]. PCA has had numerous applications in Archaeology [e.g. @mcpherron_reduction_1994; @marreiros_lithic_2013; @mccall_behavioral_2007; @mccall_multivariate_2006] and is available through most of the statistical software packages.

Here, we applied PCA to our dataset to identify possible correlations between variables that would indicate the existence of patterns that, when compared with a priori information from each of the sites, could possibly be translated into more secure proxies for occupation duration.

All analyses and data processing were accomplished in R (version 3.3.1) [@r_core_team_r:_2013]. PCA analysis was performed using the FactoMineR package [@le_factominer:_2008]. Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our supplemental online material (SOM) at https://dx.doi.org/… To produce those files we followed the procedures described by Marwick et al. [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research. The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code use for all analysis and to produce all tables and figures. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, [for more information see @marwick_computational_2016].

@@

\newline

As a result, running PCA with the enlarged set of criteria provides a slightly different perspective on the WABI patterns. Four of the calculated PCA dimensions present eigenvalues higher than 1, explaining more than 87% of dataset variability (Table \@ref(tab:eigenvalues)). Dimensions 1 and 2, alone, explain over 58% of the variability.

\newline

@@

ShortTerm <- dataset[vars]

# Set site names as row.names
row.names(ShortTerm) <- dataset$Sites

@@

\newline

```{r eigenvalues}
